### 2023-02-02 article.tex
Improve

A recent paper explores the use of paradata to increase the impact of data,
stating that lack of paradata can be seen as 'a drastic constraint'
in the use of data and offer some suggestions to 
make paradata useful for data re(use):
paradata should be comprehensive, comprehensively documentated
and be friendly to computers \cite{huvila2022improving}.
These suggestions appear to work with code as well,
where these suggestions show up as best practices (such
as \cite{stodden2013best}), some of which
described below.
The lack of computer code is just as much a drastic constraint
in the use of a scientific experiment.
The clearest effect of its absence is, most obviously,
that a computational experiment cannot be redone 
easily, yet there are wider implications, described below, as well.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Code is more important than the paper}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The experiment described above is run by code. 
It was code that detected the relationship between the genotype
(in this case, SNP \verb|rs12133641|) 
and the phenotype (in this case, the concentration of IL6RA).
To detect this association, 
there was no fieldwork nor lab-work involved.
For this experiment, 
it has been the researcher that has written the code,
after which that code does the actual work. 
A genetic epidemiologist does not need a lab, but computational power instead.

To \emph{obtain} the raw genetic data, there may have been fieldwork and/or lab-work
involved. For example, a researcher needs to go into the field to take
samples (e.g. blood) of humans or other species. 
To analyse these samples,
a researcher needs a lab. 
Depending on the analysis, there may be a big bioinformatics step
to aggregate the measurements into usable data. 
Although this paper does not focus on the code run to 
aggragate raw and noisy measurements into usable data, the same arguments
can be made for that code as in the example: that code is 
the ground truth of how the raw data are converted into useful genetic data.

% \paragraph{The code is the paradata for the results of an experiment}

Programming code (from now on: 'code') is paradata, 
as it is data that describes how data are generated.
Code is data that is usually in the form of text, 
spread over one or more files, that describes an experiment/annalysis.
The computational experiment (from now on: 'experiment') collects the data
and then looks for patterns.
Those patterns (again, a type of data) are the results of an experiment.

The (scientific) article is metadata, as it is data that describes other data:
an article describes the
experiment; an article describes what the code does in a natural 
language, such as English.
However, English (or any other natural language) is not the best
candidate to describe how the data are collected,
as it only has a loose connection with collecting the data.
If the code and article of an experiment disagree,
it is the code that actually let the data be collected.  
Instead, an article is metadata about a research.

@@

One pseudorandom example of a mismatch between code and article
was found by me in \cite{labrecque2019interpretation}.
After the corresponding author of that paper was contacted,
the code was updated and confirmed to have the same
results as published in the paper.
This sequence of events could only have happened, 
because the authors chose to publish their code.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{There are multiple way to publish code}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

As it is the code that is the most important actor in an the experiment,
it should be treated as equally or more important than the scientific
paper describing what is does.
Regardless of this, most academic journals, 
when concerning computational experiments
only review the (usually) English description of the code
and make the review of code optional.
One notable exception is rOpenSci \cite{ram2018community},
which is an organisation that does peer-review academic code 
and facilitates the optional publication of a paper.
Although both code and paper are peer-reviewed,
most effort is spent on reviewing the code,
as can be seen in, for example, 
the \verb|mcbette| R package \cite{bilderbeek2020mcbette}.

Code is commonly published alongside of a 
paper, pasted as text in the Supplementary Materials 
(for example, in \cite{labrecque2019interpretation}).
This is the rawest form of publishing code,
as it is just text.

% \paragraph{Raw code is important}

The raw code of a computational experiment is
more important to preserve than the article describing
the experiment, even when both describe the experiment identically,
correctly and reproducibly, 
as it the code that can reproduce the results.

% \paragraph{Raw code is an indicator of quality}

@@

% \paragraph{The problems with raw code}

There are multiple problems with publishing raw code.
First, there is no guarantee that that code \emph{can} actually
run, for example, due to a typo added after the code was pasted.
Secondly, there is no way to determine if code that worked yesterday
works today. These problems can be prevented by, instead of
pasting raw code, using a website to host that code (see below).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Publishing hosted code}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Code is sometimes published alongside a paper, 
in the form of a code hosting website.
This way of publishing code gives additional options,
both for the authors of that code, as well as its users.
This practice has increased,
with the abstracts from articles in the academic journal
'Bioinformatics' referring to such a hosted website
has increased tenfold, going from below 5 percent in 2009 
to around 50 percent in 2017 \cite{russell2018large}.

>>>>

A recent paper explores the use of paradata to increase the impact of data,
stating that lack of paradata can be seen as 'a drastic constraint'
in the use of data and offer some suggestions to 
make paradata useful for data re(use):
paradata should be comprehensive, comprehensively documented
and friendly to computers \cite{huvila2022improving}.
These suggestions appear to work with code as well,
where these suggestions show up as best practices (such
as \cite{stodden2013best}), some of which
described below.
The lack of computer code is just as much 'a drastic constraint'
in the use of a scientific experiment.
The clearest effect of its absence is, most obviously,
that a computational experiment cannot be redone 
easily, yet there are wider implications, described below, as well.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Code is just as important as the paper}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The experiment described above is run by code. 
It was code that detected the relationship between the genotype
(in this case, SNP \verb|rs12133641|) 
and the phenotype (in this case, the concentration of IL6RA).

The code for this experiment is vital paradata.
Like all paradata, its omission would be a constraint in the
use of the conclusion reached by it:
we can use it to assess the trustworthiness of the finding,
for example, by running the code again.
Ideally, the code is comprehensive, comprehensively documented
and be friendly to computers \cite{huvila2022improving}.
Code being friendly to computer seems like an obvious 'yes',
yet this is naive, as commonly it is hard to run code in practice. 
% TODO: Link to correct paragraphs
For code to be called computer-friendly, it must be run on an independent entity, 
as is discussed in the paragraphs about hosting code and/or putting the code in a container.
% TODO: Link to correct paragraphs
Also note that there is a problem due to the sensitivity of the data,
which is discussed in paragraph.
Lastly, note that the researcher used data that was itself generated by code:
complex bioinformatics pipelines are used to obtain reliable DNA sequences 
from the biological sample.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Making code useful paradata}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Again, paradata should be comprehensive, comprehensively documented
and friendly to computers \cite{huvila2022improving}.
In this section, these ideal properties of paradata are applied to code.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Code must be extensively documented}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Code must be extensive}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Code must be reviewed
As it is the code that is the most important actor in an the experiment,
it should be treated as equally or more important than the scientific
paper describing what is does.
Regardless of this, most academic journals, 
when concerning computational experiments
only review the (usually) English description of the code
and make the review of code optional.
One notable exception is rOpenSci \cite{ram2018community},
which is an organisation that does peer-review academic code 
and facilitates the optional publication of a paper.
Although both code and paper are peer-reviewed,
most effort is spent on reviewing the code,
as can be seen in, for example, 
the \verb|mcbette| R package \cite{bilderbeek2020mcbette}.

% Code must be clean
% \paragraph{Raw code is an indicator of quality}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Code must be computer friendly}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

% \paragraph{Code must be hosted to be moderately computer-friendly}
Code is sometimes published alongside a paper, 
in the form of a code hosting website.
This way of publishing code gives additional options,
both for the authors of that code, as well as its users.
This practice has increased,
with the abstracts from articles in the academic journal
'Bioinformatics' referring to such a hosted website
has increased tenfold, going from below 5 percent in 2009 
to around 50 percent in 2017 \cite{russell2018large}.
